---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/poissonconsulting/datacheckr.svg?branch=master)](https://travis-ci.org/poissonconsulting/datacheckr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/poissonconsulting/datacheckr?branch=master&svg=true)](https://ci.appveyor.com/project/poissonconsulting/datacheckr)
[![codecov](https://codecov.io/gh/poissonconsulting/datacheckr/branch/master/graph/badge.svg)](https://codecov.io/gh/poissonconsulting/datacheckr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/datacheckr)](https://cran.r-project.org/package=datacheckr)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/datacheckr)](https://hadley.shinyapps.io/cran-downloads)

# datacheckr

## Introduction

`datacheckr` is an R package to check data frame's rows, column names,
    column classes, values, unique keys and joins.

### Why Another Data Checking Package?

There are several existing R packages for checking data frames including
`assertr`, `assertive` and `datacheck`. They are 
great for checking data in scripts but they have several limitations 
when embedded in functions in packages. 

#### Informative Error Messages

Consider the following code.
```{r, error=TRUE}
library(assertr)
assert(mtcars, within_bounds(0,1), mpg)
```
The error message is not that helpful for a user who is not familiar with the internals of a function that has just thrown that error.

The same test using the `datacheckr::check_data()` function produces an error message 
which is more likely to allow the end user to diagnose the problem.
```{r, error = TRUE}
library(datacheckr)
check_data(mtcars, list(mpg = c(0,1)))
```

#### Intuitive Checks

Consider the data frame `data1`
```{r}
data1 <- data.frame(
  Count = c(0L, 3L, 3L, 0L), 
  LocationX = c(2000, NA, 2001, NA), 
  Extra = TRUE)
```

The following `datacheckr` code states that `data1` 
should have a column `Count` of non-missing integers with values of 0, 1 or 3, 
should not have a column `Comments` and 
*can* include a column `LocationX` with missing values between 1012 and 2345.
```{r}
check_data(data1, list(
  Count = c(0L, 1L, 3L), 
  Comments = NULL, 
  LocationX = c(NA, 2345, 1012),
  LocationX = NULL))
```

To produce similar functionality with `assertr` would require something like
(please file an [issue](https://github.com/poissonconsulting/datacheckr/issues) 
if the code below can be improved)
```{r, results='hide'}
library(magrittr) # for the piping operator
data1 %>% assert(in_set(0, 1, 3), Count) %>%
  assert_rows(num_row_NAs, within_bounds(0,0.1), Count)
stopifnot(!"Comments" %in% colnames(data1))
if ("LocationX" %in% colnames(data1))
  data1 %>% assert(within_bounds(1012, 2345), LocationX)
```
which is in my opinion less intuitive.

#### A Single Function Call

The above checks can be performed
on several data frames by simply repeatedly calling `check_data()`
```{r}
data3 <- data2 <- data1

values <- list(
  Count = c(0L, 1L, 3L), 
  Comments = NULL, 
  LocationX = c(NA, 2345, 1012),
  LocationX = NULL)

check_data(data1, values)
check_data(data2, values)
check_data(data3, values)
```

The same tests using `assertr` would require the 
`assertr` code above to
be copied and pasted three times
which is tedious to produce and read; and as a result error prone.

## Installation

To install the release version from CRAN
```
install.packages("datacheckr")
```

Or the development version from GitHub
```
# install.packages("devtools")
devtools::install_github("poissonconsulting/datacheckr")
```

## Contribution

Please report any [issues](https://github.com/poissonconsulting/datacheckr/issues).

[Pull requests](https://github.com/poissonconsulting/datacheckr/pulls) are always welcome.

